---
title: The Shiny Variant Explorer
author:
    -   name: Kévin Rue-Albrecht
        email: k.rue-albrecht@imperial.ac.uk
        affiliation: Department of Medicine, Imperial College London,
            Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
    RStudio/Shiny web-application demonstrating the functionalities of
    the *TVTB* package integrated in a programming-free environment.
vignette: >
    %\VignetteIndexEntry{The Shiny Variant Explorer}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
output: 
    BiocStyle::pdf_document2:
        toc_newpage: true
bibliography:
    TVTB.bib
---

# Preliminary notes {#Preliminary}

The Shiny Variant Explorer (*tSVE*) was primarily developped to demonstrate
features implemented in the `r Biocpkg("TVTB")`.
As a result, a few important considerations should be made
to clarify what should and should **not** be expected from the 
web-application:

* It is technically not feasible to offer in a web-interface the same degree
    of flexibility as the command-line environment (*e.g.* `...`[^1]).
* Greater control over the input and output data is possible at the
    command-line (*e.g.* refinement `ggplot` objects,
    definition of custom genomic ranges)
* Requests for new feature should apply to the package first.
    Only features relevant to the package functionalities
    may be made available in the web-application.
* Figures (`ggplot`) are currently the only output that can be exported from
    the web-application (using the web browser
    "Download image", "Copy image", or equivalent context menu item).
    In the future, action buttons may be added to export tables (`data.frame`).
* This vignette is largely static as the web-application may only be used
    in an interactive session.
* First-time users are encouraged to follow this vignette
    sequentially (*i.e* without skipping sections),
    as it takes users through the sequence of actions of a typical analysis
    + Actions are indicated by the word **Action** followed by
        indented bullet points

[^1]: The `...` argument is called "ellipsis".

# Pre-requisites {#Prerequisites}

The *Shiny Variant Explorer* suggests a few additional package dependencies
relative to the package, to support relevant forms of data input and display.

**Input**

* The `r Biocpkg("ensembldb")` package and relevant `EnsDb`[^2] annotation
    package are required if using that interface to query genomic ranges
* The `r Biocpkg("rtracklayer")` package is required if a BED file is used
    to provide genomic ranges

[^2]: In the future, the web-application may also
    support `TxDb` and `OrganismDb` annotation packages.

Certain [sections](#InputEnsDb) of this vignette require that the
`r Biocpkg("EnsDb.Hsapiens.v75")` package be installed.
This can be done with the following chunk of code:

```{r EnsDb.Hsapiens.v75, message=FALSE}
if (!require(EnsDb.Hsapiens.v75, quietly = TRUE))
    biocLite("EnsDb.Hsapiens.v75")
```

**Display**

* The latest version of the `r Githubpkg("rstudio/DT")` package is recommended
    to benefit from the latest developments (*e.g.* column filters inactivated
    if a single value exist in that column; version *>= 0.2.2*)
* The `r Githubpkg("rstudio/shiny")` package is required for all Shiny
    web-applications

# Launching the Shiny Variant Explorer {#Launch}

The `TVTB::tSVE()` method launches the web-application.

# Overall layout of the web-application {#OverallLayout}

The web-application is generally implemented as a web-page
with a top level navigation bar organised 
from left to right to reflect progression through a typicaly analysis,
with the exception of the last two menu items **Settings** and **Session**,
which may be useful to check and update at any point.

Here is a brief overview of the menu items:

* **Input**
    + Control the samples, phenotypes, genomic ranges, and VCF fields
        imported.
    + An `EnsDb` annotation package may be selected to use the associated
        database interface.
* **Frequencies**
    + Add and remove INFO fields that contain calculated genotype counts
        and allele frequencies.
    + Counts and frequencies can be obtained overall (*i.e.* across all
        samples), or within individual phenotype levels.
* **Filters**
    + Define and apply *VCF filter rules* (detailed in a separate vignette).
* **Views**
    + Display and examine major objects of the analysis and their slots
* **Plots**
    + Display and refine data plots
* **Settings**
    + Control advanced parameters of the analysis
* **Session**
    + Display session information and other relevant information

# Input panel {#Input}

The **Input** panel controls the major input parameters of the analysis,
including phenotypes (and therefore samples), genomic ranges, and fields
to import from VCF file(s).
Those inputs are useful to import only data of interest, as well as
to limit memory usage and duration of the calculations.

## Phenotypes {#InputPhenotypes}

Phenotypes are critical to define groups of samples that may be compared
in summary statistics, tables, and plots.
Moreover, phenotypes also implicitely define the set of samples required
in the analysis (usually including unique sample identifiers as `rownames`).

The web-application accepts phenotypes stored in text files, with the
following requirements:

* Fields must be delimited by "white space" (default of `read.table`)
* The first column of the file must contain unique sample identifiers
* The first row of the file must phenotype names, as syntactically valid
    `colnames`

When provided, phenotypes will be used to import from VCF file(s)
only genotypes for the corresponding samples identifiers.
Moreover, an error will be raised if any of the sample identifiers present
in the phenotypes is absent from the VCF file(s).

Note that the web-application does not *absolutely* require phenotype
information. In the absence of phenotype information, all samples are imported
from VCF file(s).

> **Action**:
> 
> * Click on the *Browse* action button
> * Navigate to the *extdata* folder of the *TVTB* installation directory
> * Select the file *integrated_samples.txt*

**Notes**

* The `r Biocpkg("TVTB")` installation directory can be identified
using the following command in an *R* session:

```{r systemFile, results='hide'}
system.file("extdata", package = "TVTB")
```

* The file selection window that is open by the action button browses
    files on the server side.
    + This is only relevant if the package/web-application is run on a remote
        server
    + If the web-application is run locally, this does not make any difference

## Genomic ranges {#InputGRanges}

Genomic ranges are critical to import only variants in
targeted genomic regions or features (*e.g.* genes, transcripts, exons),
as well as to limit memory usage and duration of calculations.

The Shiny Variant Explorer currently supports three types of input to define
genomic ranges:

* BED file
* UCSC-style text field
* `EnsDb` annotation packages

Currently, the web-application uses genomic ranges solely
to query the corresponding variants from VCF file(s).
In the future, those genomic ranges may also be used to produce
faceted summary statistics and plots.

**Notes**:

* The web-application does not *absolutely* require genomic ranges.
In the absence of genomic ranges, all variants are imported from VCF file(s).
*Use with caution!*
* The active genomic ranges are only taken from the currently selected input
    mode

### BED file {#InputBED}

If a BED file is supplied, the web-application parses it using the
`rtracklayer::import.bed` method.
Therefore the file must respect the
[BED file format](https://genome.ucsc.edu/FAQ/FAQformat.html#format1)
guidelines.

> **Action**:
> 
> * Click on the *Browse* action button
> * Navigate to the *extdata* folder of the *TVTB* installation directory
> * Select the file *SLC24A5.bed*

**Notes**:

* The BED file defines the same genomic range
    that was used to extract variant from the 
    [1000 Genomes Project](http://www.1000genomes.org) Phase 3
    release VCF file used in this vignette (see section
    [Variants](#InputVariants)).
* The file selection window that is open by the action button browses
    files on the server side (see [Phenotypes](#InputPhenotypes) section
    above).

### UCSC format {#InputUCSC}

Sequence names (*i.e.* chromosomes), start, and end positions of one or more
genomic ranges may be defined in the text field,
with individual regions separated by `";"`.

> **Action**:
> 
> * Paste `15:48413169-48434869` in the text field

### Ensembl-based annotation packages {#InputEnsDb}

Currently, genomic ranges encoding only gene-coding regions may be retrieved
from an Ensembl-based database.
This feature was adapted from the web-application implemented in the
`r Biocpkg("ensembldb")` package.

\bioccomment{
In the future, the interface to query transcript and exons annotations
may be added to the web-application.
}

\fixme{
Genomic feature located on contigs may cause problems when working
with one VCF per chromosome.
In the future, an option may be added to ignore contigs.
}

> **Action**:
> 
> * Paste `SLC24A5` in the text field

## Variants {#InputVariants}

At the core of the `r Biocpkg("TVTB")` package, variants must be imported from
one or more VCF file(s), optionally annotated by the
[Ensembl Variant Effect Predictor](http://www.ensembl.org/info/docs/tools/vep)
(VEP)[@RN1].

Considering the large size of most VCF file(s), it is common practice to
split genetic variants into multiple files, each containing variants
in a single chromosome.
The Shiny Variant Explorer supports two situations:

* all variants are stored in a single VCF file ("Single-VCF" mode)
* variants are split into one file per sequence, with the requirement that
    files be named with a pattern including the sequence name
    (identical to the `seqnames` slot of the genomic ranges described
    [above](#InputGRanges) ("Multi-VCF mode").

In addition, VCF files can store a plethora of information in their various
fields. It is often useful to select only a subset of fields relevant for
a particular analysis, to limit memory usage during an analysis.
The application uses the
`r Biocpkg("VariantAnnotation")` `scanVcfHeader` to parse the header of
the VCF file (*Single-VCF* mode) or the first VCF file (*Multi-VCF* mode),
to display the list of available fields that users may choose to import.
A few considerations must be made:

* The web-application requires that Ensembl VEP predictions be present in the
    INFO field
* The web-application requires that the `"GT"` key be present in the
    FORMAT field

### Single-VCF mode {#SingleVCF}

This mode display an action button that must be used to select the VCF file
from which to import variants.

> **Action**:
> 
> * Click on the *Browse* action button
> * Navigate to the *extdata* folder of the *TVTB* installation directory
> * Select the file *chr15.phase3_integrated.vcf.gz*

### Multi-VCF mode {#MultiVCF}

This mode requires two pieces of information:

* The path to the folder that contains one or more VCF file(s)
* The naming pattern of VCF file(s), with the following requirement:
    + The pattern must include `"%s"` to declare the emplacement of the
        sequence (*i.e.* chromosome) name in the pattern

Note that a summary of VCF file(s) detected using the given the folder and
pattern is displayed on the right, to help users determine whether the
parameters are correct. In addition, the content of the given folder is
displayed at the bottom of the page, beside the same content filtered for
the VCF file naming pattern.

> **Action**:
> 
> *None*. The text fields should already be filled with default values,
> pointing to the single example VCF file (a subset of chromosome 15).

### VCF scan parameters {#scanVcfParam}

This panel allows users to select the INFO and FORMAT fields to import
(in the `info` and `geno` slots of the `VCF` object, respectively).

It is important to note that the `FORMAT/GT` and `INFO/<vep>` fields---where
`<vep>` stands for the INFO key where Ensembl VEP predictions are stored---are
implicetly imported from the VCF.
Similarly, the mandatory FIXED fields `CHROM`, `POS`, `ID`, `REF`, `ALT`,
`QUAL`, and `FILTER` are automatically imported to populate
the `rowRanges` slot of the `VCF` object.

> **Action**:
> 
> * Click the *Deselect all* action button under the **INFO fields** 
>     selection input to import only the INFO/CSQ and FORMAT/GT fields.
> * Click the *Import variants* action button
>
> A summary of variants, phenotypes, and samples imported will appear beside
> the action button.

## EnsDb Annotation packages {#EnsDbPkg}

This panel allows users to select an pre-installed `EnsDb` annotation package.
Currently, annotation packages are only used to query gene-coding regions
which are then used to import from VCF file(s) variants
only in those genomic ranges.

> **Action**:
> 
> * If none of the `EnsDb` packages are installed, it will simply not be
>     possible to use the `ensembl` interface of the *Genomic ranges* input
>     tab.
> * If the `EnsDb.Hsapiens.v75` package is the only `EnsDb` packages installed,
>     no action is required; the package should already be pre-selected.
> * If the `EnsDb.Hsapiens.v75` package is **not** the only `EnsDb` packages
>     installed, users should select it in the list of choices.

# Frequencies panel {#Frequencies}

This panel demonstrates the use of three methods implemented in the
`r Biocpkg("TVTB")` package, namely `addFrequencies`, `addOverallFrequencies`,
and `addPhenoLevelFrequencies`.

## Overall frequencies {#OverallFrequencies}

This panel allows users to *Add* and *Remove* INFO fields
that contain genotype counts
(*i.e.* homozygote reference, heterozygote, homozygote alternate)
and allele frequencies
(*i.e.* alternate allele frequency, minor allele frequency)
calculated across all the samples and variants imported.
The web-application uses the homozygote reference, heterozygote,
and homozygote alternate genotypes defined in the **Advanced settings**
panel.

Importantly, the name of the INFO keys that are used to store
the calculated values can be defined in the **Advanced settings** panel.

> **Action**:
> 
> * Click the *Add* action button
> * See the *Latest changes* message update at the top of the screen.

\fixme{
Add cross-references to the "Advanced settings" section when it is written.
}

## Phenotype-level frequencies {#PhenoLevelFrequencies}

This panel allows users to *Refresh* the list of INFO fields
that contain genotype counts and allele frequencies
calculated within groups of samples associated with
various levels of a given phenotype.

> **Action**:
> 
> * Select *super_pop* in the list of phenotypes
> * Click the *Select all* action button
> * Click the *Refresh* action button
> * See the *Latest changes* message update at the top of the screen.

# VCF filter rules {#VcfFilterRules}

One of the flagship features of the `r Biocpkg("TVTB")` package
are the *VCF filter rules*, extending the
`r Biocpkg("S4Vectors")` `FilterRules` class to new classes of filter rules
that can be evaluated within environments defined by the various slots
of `VCF` objects.

Generally speaking, `FilterRules` greatly facilitate the design 
and combination of powerful filter rules for table-like objects,
such as the `fixed` and `info` slots of
`r Biocpkg("VariantAnnotation")` `VCF` objects,
as well as Ensembl VEP predictions stored as `GRanges` returned by the
`r Biocpkg("ensemblVEP")` `parseCSQToGRanges` method.

A separate vignette described in greater detail the use of classes
that contain *VCF filter rules*. A simple example is shown below.

> **Action**:
> 
> * Select *VEP* as the *Type* of filter
> * Paste `grepl("missense",Consequence)` in the text field
> * Leave the *Active?* checkbox ticked
> * Click the *Add filter* action button
> * See the list of rules update at the bottom of the screen
> * Click the *Apply filters* action button
> * See the summary of filtered variants update beside the action button

# Views {#Views}

This panel offers the chance to examine the main objects of the session,
namely:

* the active genomic ranges
* `rowRanges` and selected meta-columns of the filtered variants
* selected field of the `info` slot (of the filtered variants)
* selected Ensembl VEP predictions (of the filtered variants)
* selected phenotypes attached to the variants
* subset of genotypes (among the filtered variants)
    + genotypes of all filtered variants may be displayed as a heatmap
        (`ggplot`)
        
> **Action**:
> 
> * In the various panels, select fields to examine each object
> * In the *Heatmap* tab of the *Genotypes* panel
> * Click the *Go!* action button to calculate and display the heatmap

# Plots {#Plots}

This panel demonstrates the use of two methods implemented in the
`r Biocpkg("TVTB")` package, namely `tabulateVepByPhenotype`
and `densityVepByPhenotype`.

## Count and tabulate discrete VEP predictions by phenotype level {#TVBP}

The `tabulateVepByPhenotype` method is designed to count 
Ensembl VEP predictions associated with variants observed in the distinct
levels of a given phenotype.
The method can display those counts in two different ways:

* a `r CRANpkg("ggplot")` `geom_barplot`
* a `data.frame` in wide format

Typically, those counts may be faceted by genomic feature (*i.e.* transcript),
as the VEP script may produce one prediction per *Variant : Feature*
combination. In other words, faceting counts by feature ensures that
each variant contributes at most one count to each feature.

> **Action**:
> 
> * Select *Consequence* as the *Variant effect prediction*
> * Select *super_pop* as the *Phenotype field*
> * Select *Feature* as the *VEP faceting key*
> * Click the *Advanced controls* checkbox
> * Decrease the *Legend font size* to `0.6`
> * Rotate the *Angle of X labels* to `90` degrees
> * Click the *Apply* action button to calculate and display the barplot
> * Hover the plot with the cursor to display useful information!

In addition, the web-application also displays the counts of VEP predictions
in long format in the *Table* panel, more convenient than the wide format
for filtering.

> **Action**:
> 
> * Use the filtering fields at the top of the table to examine the counts
>     (e.g. *missense_variant* and *ENST00000341459*)

## Distribution of continuous VEP predictions by phenotype level {#DVBP}

The `densityVepByPhenotype` method is designed to parse continuous 
Ensembl VEP predictions associated with variants observed in the distinct
levels of a given phenotype, and display the distribution of those predictions
in those phenotype levels.

The method can return this information in two different ways:

* a `r CRANpkg("ggplot")` object with `geom_density` and/or `geom_dotplot`
    layers
* a `data.frame` of tabulated values in long format

As described above for [discrete VEP predictions](#TVBP), it may be useful
to tabulate continuous VEP predictions by genomic feature to ensure that
each variant contributes at most once to each faceted plot.

> **Action**:
> 
> * Select or type *CADD_PHRED* (notice the autocompletion feature)
    as the *Variant effect prediction*
> * Select *super_pop* as the *Phenotype field*
> * Leave *Density & Dot plot* as the *Layer(s)*
> * Select *Feature* as the *VEP faceting key*
> * Click the *Apply* action button to calculate and display the `ggplot`
> * See how African and East Asian population tend to display more damaging
>     mutations
> 
> Optionally:
> * Click the *Advanced controls* checkbox
> * Decrease the *Relative size of X text* and *Relative size of Y text* to
>     `0.7`
> * Click the *Apply* action button to refresh the `ggplot`

# Session info {#SessionInfo}

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo}
sessionInfo()
```

[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
