---
title: VCF filter rules
author:
    -   name: Kévin Rue-Albrecht
        email: k.rue-albrecht@imperial.ac.uk
        affiliation: Department of Medicine, Imperial College London,
            Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('TVTB')`"
abstract: >
    Currently, `VCF` objects of the `r Biocpkg("VariantAnnotation")` package
    may be subsetted by indices or names of variant records.
    The `r Biocpkg("TVTB")` package extends `FilterRules` of the
    `r Biocpkg("S4Vectors")` package to provide filter rules applicable to
    `VCF` objects. These new classes of filter rules provide containers
    for powerful expressions and functions to facilitate filtering of variants
    using combinations of filters applicable to FIXED fields, INFO fields,
    and Ensembl VEP predictions stored in a given INFO field of VCF files.
vignette: >
    %\VignetteIndexEntry{VCF filter rules}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
output:
    BiocStyle::pdf_document2:
        toc_newpage: true
bibliography:
    TVTB.bib
---

# Motivation {#Motivation}

## Background

`VCF` objects of the `r Biocpkg("VariantAnnotation")` package contain a
plethora of information imported from specific fields of source VCF files and
stored in dedicated slots (*e.g.* `fixed`, `info`, `geno`),
as well as optional Ensembl VEP predictions [@RN1]
stored under a given key of their INFO slot.

This information may be used to identify and filter variants of interest
for further analysis.
However, the size of genetic data sets and the variety of filter
rules---and their combinatorial explosion---create
considerable challenges in terms of workspace memory and entropy
(*i.e.* size and number of objects in the workspace, respectively).

The `FilterRules` class implemented in the `r Biocpkg("S4Vectors")` package
provides a powerful tool to create flexible 
and lightweight filter rules defined in the form of
`expression` and `function` objects that can be evaluated
within given environments.
The `r Biocpkg("TVTB")` package extends this `FilterRules` class
into novel classes of *VCF filter rules*,
applicable to information stored in the distinct slots of `VCF`
objects (*i.e.* `CollapsedVCF` and `ExpandedVCF` classes),
as described below:

|       Class       | Motivation                                              |
| :--------------:  | :------------------------------------------------------ |
| `VcfFixedRules`   | Filter rules applicable to the `fixed` slot of a        |
|                   | `VCF` object.                                           |
|                   |                                                         |
|  `VcfInfoRules`   | Filter rules applicable to the `info` slot of a         |
|                   | `VCF` object.                                           |
|                   |                                                         |
|  `VcfVepRules`    | Filter rules applicable to the Ensembl VEP predictions  |
|                   | stored in a given INFO key of a `VCF` object.          |
|                   |                                                         |
| `VcfFilterRules`  | Combination of `VcfFixedRules`, `VcfInfoRules`, and     |
|                   |  `VcfVepRules` applicable to a `VCF` object.            |

Table: Motivation for each of the new classes extending `FilterRules`
to define *VCF filter rules*.

Note that `FilterRules` objects themselves are applicable to `VCF` objects,
with two important difference from the above specialised classes:

* Expressions must explicitely refer to the `VCF` slots
* As a consequence, a single expression can refer to fields from different
    `VCF` slots, for instance:

```{r FilterRules}
fr <- S4Vectors::FilterRules(list(
    mixed = function(x){
        VariantAnnotation::fixed(x)[,"FILTER"] == "PASS" &
            VariantAnnotation::info(x)[,"MAF"] >= 0.05
    }
))
fr
```

## Features {#Features}

As they inherit from the `FilterRules` class,
these new classes benefit from accessors and methods defined for their
parent class, including:

* *VCF filter rules* can be toggled individually between an active and
    an inactive states
* *VCF filter rules* can be subsetted, edited, replaced, and deleted

To account for the more complex structure of `VCF` objects, some of the new
*VCF filter rules* classes implemented in the `r Biocpkg("TVTB")` package
require additional information stored in new dedicated slots,
associated with the appropriate accessors and setters.
For instance:

* `VcfVepRules` require the INFO key where predictions of the
    Ensembl Variant Effect Predictor are stored in a `VCF` object.
    The `vep` accessor method may be used to access this slot.
* `VcfFilterRules`---which may combine any number of filter rules stored in
    `FixedRules`, `VcfFixedRules`, `VcfInfoRules`, `VcfVepRules`, and other
    `VcfFilterRules` objects---
    mark each filter rule with their type in the combined
    object. The information is stored in the `type` slot, which may be
    accessed using the *read-only* accessor method `type`.

# Demonstration data {#DemoData}

For the purpose of demonstrating the utility and usage of *VCF filter rules*,
a set of variants and associated phenotype information was
obtained from the
[1000 Genomes Project](http://www.1000genomes.org) Phase 3 release.
It can be imported as a `CollapsedVCF` object using the following code:

```{r makeCollapsedVCF}
library(TVTB)
extdata <- system.file("extdata", package = "TVTB")
vcfFile <- file.path(extdata, "chr15.phase3_integrated.vcf.gz")
tabixVcf <- Rsamtools::TabixFile(file = vcfFile)
vcf <- VariantAnnotation::readVcf(file = tabixVcf)
```

VCF filter rules may be applied to `ExpandedVCF` objects equally:

```{r makeExpandedVCF}
evcf <- VariantAnnotation::expand(x = vcf, row.names = TRUE)
```

## CollapsedVCF and ExpandedVCF {#VcfClasses}

As described in the documentation of the `r Biocpkg("VariantAnnotation")`
package, the key difference between `CollapsedVCF` and `ExpandedVCF` objects
---both extending the `VCF` class---is
the expansion of multi-allelic records into bi-allelic records, respectively.
In other words (quoting the `r Biocpkg("VariantAnnotation")` documentation):

> "`CollapsedVCF` objects contains the ALT data as a
> `DNAStringSetList` allowing for multiple alleles per variant.
> In contrast, the `ExpandedVCF` stores the ALT data as a `DNAStringSet`
> where the ALT column has been expanded to create a flat form of the data
> with one row per variant-allele combination."

This difference has implications for filter rules using the `"ALT"` field
of the `info` slot, as demonstrated in a [later section](#AltField).

# Fields available for the definition of filter rules {#VcfFields}

First, let us examine which fields (*i.e.* column names)
are available in the `VCF` objects to create *VCF filter rules*:

```{r simpleRules}
fixedVcf <- colnames(fixed(vcf))
fixedVcf
infoVcf <- colnames(info(vcf))
infoVcf
csq <- ensemblVEP::parseCSQToGRanges(x = evcf)
vepVcf <- colnames(mcols(csq))
vepVcf
```

\bioccomment{
Although multi-allelic records present in `CollapsedVCF` objects 
are expanded into bi-allelic records
in `ExpandedVCF` objects, names fields in the different slots are identical
in the corresponding collapsed and expanded objects.
}

# Usage of VCF filter rules {#FilterUsage}

## Filter rules using a single field {#SingleFieldFilter}

The value of a particular field can be used to define expressions
that represent simple filter rules based on that value alone.
Multiple rules may be stored in any one `FilterRules` objects.
Ideally, *VCF filter rules* should be named
to facilitate their use,
but also as a reminder of the purpose of each particular rule.
For instance, in the chunk of code below, two filter rules are defined
using fields of the `fixed` slot:

* A rule named `"pass"` identifies variants for which the value in the
    `FILTER` field is `"PASS"`
* A rule named `"qual20"` identifies variants where the value in the `QUAL`
    field is greater than or equal to `20`

```{r simpleFixedRules}
fixedRules <- VcfFixedRules(exprs = list(
    pass = expression(FILTER == "PASS"),
    qual20 = expression(QUAL >= 20)
))
active(fixedRules)["qual20"] <- FALSE
summary(evalSeparately(fixedRules, vcf))
```

In the example above, all variants pass the active `"pass"` filter,
while the deactivated rules `"qual20"`) automatically returns `TRUE`
for all variants.

\bioccomment{
`active(fixedRules["qual20"]) <- FALSE` would be wrong syntax, as it
does not change the active state of the `fixedRules` object;
instead, it changes the active state of a temporary object made of the
selected filter rule. As a consequence, the `fixedRules` object would be
left unchanged after this command.
}

## Filter rules using multiple fields {#MultipleFieldsFilter}

It is also possible for *VCF filter rules* to use multiple fields
(of the same `VCF` slot) in a single expression.
In the chunk of code below, the *VCF filter rule* identifies variants
for which both the `"REF"` and `"ALT"` values (in the INFO slot)
are one of the four nucleotides
(*i.e.* a simple definition of single nucleotide polymorphisms; SNPs):

```{r addFixedRule}
nucleotides <- c("A", "T", "G", "C")
SNPrule <- VcfFixedRules(exprs = list(
    SNP = expression(
    as.character(REF) %in% nucleotides &
        as.character(ALT) %in% nucleotides)
))
summary(evalSeparately(SNPrule, evcf, enclos = .GlobalEnv))
```

Some considerations regarding the above filter rule:

* considering that the filter rule requires the `nucleotides` character vector,
    the global environment must be supplied as the enclosing environment to
    successfully evaluate the expression
* `"REF"` and `"ALT"` are stored as `DNAStringSet` in `CollapsedVCF` objects
    and must be converted to `character`
    in order to successfully apply the method `%in%`.

## Calculations in filter rules {#CalculationFilter}

Expressions that define filter rules may also include calculations.
In the chunk of code below, two simple *VCF filter rules* are defined
using fields of the `info` slot:

* A rule named `"samples"` identifies variants where at least 90% of samples
    have data (*i.e.* the `NS` value is greater than or equal to `0.9`
    times the total number of samples)
* A rule named `"avgSuperPopAF"` calculates the average of the allele
    frequencies calculated in each the five super-populations
    (available in several INFO fields), and
    subsequently identifies variants with an average value greater than `0.05`.

```{r simpleInfoRules}
infoRules <- VcfInfoRules(exprs = list(
    samples = expression(NS > (0.9 * ncol(evcf))),
    avgSuperPopAF = expression(
        (EAS_AF + EUR_AF + AFR_AF + AMR_AF + SAS_AF) / 5 > 0.05
    )
))
summary(evalSeparately(infoRules, evcf, enclos = .GlobalEnv))
```

\bioccomment{
Here, the global environment must be supplied as the filter rule requires
the `evcf` object itself to access its number of columns
(not only as the environment in which the rules must be evaluated).
}

## Functions in filter rules {#FunctionFilter}

It may be more convenient to define filters as a `function` objects.
For instance, the chunk of code below:

* first, defines a function that:
    + expects the `info` slot of a `VCF` object as input
    + identifies variants where at least two thirds of the super-populations
        show an allele frequency greater than 5%
* next, defines a *VCF filter rule* using the above function

```{r infoFunctionRule}
AFcutoff <- 0.05
popCutoff <- 2/3
filterFUN <- function(envir){
    # info(vcf) returns a DataFrame; rowSums below requires a data.frame
    df <- as.data.frame(envir)
    # Identify fields storing allele frequency in super-populations
    popFreqCols <- grep("[[:alpha:]]{3}_AF", colnames(df))
    # Count how many super-population have an allele freq above the cutoff
    popCount <- rowSums(df[,popFreqCols] > AFcutoff)
    # Convert the cutoff ratio to a integer count
    popCutOff <- popCutoff * length(popFreqCols)
    # Identifies variants where enough super-population pass the cutoff
    testRes <- (popCount > popCutOff)
    # Return a boolean vector, required by the eval method
    return(testRes)
}
funFilter <- VcfInfoRules(exprs = list(
    commonSuperPops = filterFUN
))
summary(evalSeparately(funFilter, evcf))
```

\bioccomment{
Filter rules written as functions may use values in the the global environment,
without the need to supply it as the enclosing environment.
}

Notably, the `filterFUN` function may also be applied separately to the
`info` slot of `VCF` objects:

```{r applyInfoFunction}
summary(filterFUN(info(evcf)))
```

## Pattern matching in filter rules {#PatternFilter}

The `grepl` function is particularly suited for the purpose of `FilterRules`
as they return a `logical` vector:

```{r greplFilter}
missenseFilter <- VcfVepRules(
    exprs = list(
        exact = expression(Consequence == "missense_variant"),
        grepl = expression(grepl("missense", Consequence))
        ),
    vep = "CSQ")
summary(evalSeparately(missenseFilter, evcf))
```

In the above chunk of code:

* the filter rule named `"exact"` matches only the given value, associated
    with `27` variants,
* the filter rule named `"grepl"` also matches an extra two variants
    associated with the value `"missense_variant&splice_region_variant"`
    matched by the given pattern. By deduction, the two rules indicate
    together that those two variants were not assigned the
    `"missense_variant"` prediction.

# Using ALT data in the `fixed` slot of `VCF` objects {#AltField}

As detailed in an earlier section introducing the
[demonstration data](#VcfClasses), and more thoroughly in the documentation of
the `r Biocpkg("VariantAnnotation")` package, 
`CollapsedVCF` and `ExpandedVCF` classes differ in the class of data
stored in the `"ALT"` field of their respective `fixed` slot.
As as result, *VCF filter rules* using
data from this field must take into account the `VCF` class in order to
handle the data appropriately:

## `ExpandedVCF` objects {#ALTExpandedVCF}

A key aspect of `ExpandedVCF` objects is that the `"ALT"` field of their
`fixed` slot may store only a single allele per record as a `DNAStringSet`
object.

For instance, in an earlier section that demonstrated
[Filter rules using multiple raw fields](#MultipleFieldsFilter),
ALT data of the `fixed` slot in an `ExpandedVCF` object had to be re-typed
from `DNAStringSet` to `character` before the `%in%` function could be applied.

Nevertheless, *VCF filter rules* may also make use of methods associated with
the `DNAStringSet` class.
For instance, genetic insertions may be identified
using the fields `"REF"` and `"ALT"` fields of the `fixed` slot:

```{r fixedInsertionFilter}
fixedInsertionFilter <- VcfFixedRules(exprs = list(
    isInsertion = expression(
        Biostrings::width(ALT) > Biostrings::width(REF)
    )
))
evcf_fixedIns <- subsetByFilter(evcf, fixedInsertionFilter)
as.data.frame(fixed(evcf_fixedIns)[,c("REF", "ALT")])
```

Here, the above `VcfFixedRules` is synonym to a distinct
`VcfVepRules` using the Ensembl VEP prediction `"VARIANT_CLASS"`:

```{r vepInsertionFilter}
vepInsertionFilter <- VcfVepRules(exprs = list(
    isInsertion = expression(VARIANT_CLASS == "insertion")
))
evcf_vepIns <- subsetByFilter(evcf, vepInsertionFilter)
as.data.frame(fixed(evcf_vepIns)[,c("REF", "ALT")])
```

## `CollapsedVCF` objects {#ALTCollapsedVCF}

In contrast to `ExpandedVCF`, `CollapsedVCF` may contain more than one allele
per record in their `"ALT"` field (`fixed` slot),
represented by a `DNAStringSetList` object.

As a result, *VCF filter rules* using the `"ALT"` field of
the `info` slot in `CollapsedVCF` objects
may use methods dedicated to `DNAStringSetList` to handle the data.
For instance,
multi-allelic variants may be identified by the following `VcfFixedRules`:

```{r fixedMultiallelicFilter}
multiallelicFilter <- VcfFixedRules(exprs = list(
    multiallelic = expression(lengths(ALT) > 1)
))
summary(eval(multiallelicFilter, vcf))
```

# Combination of multiple types of *VCF filter rules*  {#VcfFilterRules}

Any number of `VcfFixedRules`, `VcfInfoRules`, and `VcfVepRules`---or
even `VcfFilterRules` themselves---may
be combined into a larger object of class `VcfFilterRules`.
Notably, the *active* state of each filter rule is transferred to the
combined object.
Even though the `VcfFilterRules` class acts as a container for multiple types
of *VCF filter rules*, the resulting `VcfFilterRules` object
also extends the `FilterRules` class, and as a result
can be evaluated and used to subset `VCF` objects identically to any of
the specialised more specialised classes.

During the creation of `VcfFixedRules` objects, each *VCF filter rule*
being combined is marked with a *type* value,
indicating the VCF slot in which the filter rule must be evaluated.
This information is stored in the new `type` slot of `VcfFixedRules` objects.
For instance, it is possible to combine two `VcfFixedRules`
(containing two and one filter rules, respectively), one 
`VcfInfoRules`, and one `VcfVepRules` defined earlier in this vignette:

```{r combineRules}
vignetteRules <- VcfFilterRules(
    fixedRules,
    SNPrule,
    infoRules,
    vepInsertionFilter
)
vignetteRules
active(vignetteRules)
type(vignetteRules)
summary(evalSeparately(vignetteRules, evcf, enclos = .GlobalEnv))
```

Clearly[^1], the *VCF filter rules* `SNP` and `isInsertion` are
mutually exclusive, which explains the final `0` variants left after filtering.
Conveniently, either of these rules may be deactivated before evaluating the
remaining active filter rules:

[^1]: This statement below would be more evident if the `summary` method
was displaying the result of `evalSeparately` in this vignette as it does it
in an R session.

```{r combinedRulesDeactivated}
active(vignetteRules)["SNP"] <- FALSE
summary(evalSeparately(vignetteRules, evcf, enclos = .GlobalEnv))
```

As a result, the deactivated filter rule (`"SNP"`) now returns `TRUE`
for all variants, leaving a final `2` variants[^2] pass the remaining
active filter rules:

[^2]: Again, this statement would benefit from the result of `evalSeparately`
being displayed identically to an R session.

* INFO/FILTER equal to `"PASS"`
* INFO/NS greater than 90% of the number of samples in the data set
* Average of super-population allele frequencies greater than `0.05`
* Ensembl VEP prediction `VARIANT_CLASS` equal to `"insertion"`

Finally, the following chunk of code demonstrates how
`VcfFilterRules` may also be created from the
combination of `VcfFilterRules`, either with themselves
or with any of the classes that define more specific *VCF filter rules*.
Notably, when `VcfFilterRules` objects are combined,
the `type` and `active` value of each filter rule is transferred to
the combined object:

**Combine `VcfFilterRules` with `VcfVepRules`**
```{r combineFilterVep}
combinedFilters <- VcfFilterRules(
    vignetteRules, # VcfFilterRules
    missenseFilter # VcfVepRules
)
type(vignetteRules)
type(combinedFilters)
active(vignetteRules)
active(missenseFilter)
active(combinedFilters)
```

**Combine multiple `VcfFilterRules` with `VcfFilterRules` (and more)**

To demonstrate this action, another `VcfFilterRules` must first be created.
This can be achieve by simply re-typing a `VcfVepRules` defined earlier:

```{r combineFilterFilter}
secondVcfFilter <- VcfFilterRules(missenseFilter)
secondVcfFilter
```

It is now possible to combine the two `VcfFilterRules`.
Let us even combine another `VcfInfoRules` object in the same step:

```{r combineMultipleRules}
manyRules <- VcfFilterRules(
    vignetteRules, # VcfFilterRules
    secondVcfFilter, # VcfFilterRules
    funFilter # VcfInfoRules
)
manyRules
active(manyRules)
type(manyRules)
summary(evalSeparately(manyRules, evcf, enclos = .GlobalEnv))
```

Critically, users must be careful to combine rules **all** compatible with
the class of `VCF` object in which it will be evaluated
(*i.e.* `CollapsedVCF` or `ExpandedVCF`).

# Session info {#SessionInfo}

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References {#References}

[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
